---
title: "Getting Started with spatialGE"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with spatialGE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: vignette_references.bib
---

## What you'll learn

This quick-start guide will get you analyzing spatial transcriptomics data in under 10 minutes. You'll learn:

- How to load spatial transcriptomics data into an `STlist` object
- How to visualize gene expression and cluster patterns with `STplot`
- How to calculate spatial autocorrelation with `SThet`
- How to identify spatially variable genes
- Where to go next for advanced analyses

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Installation

```{r install, eval=FALSE}
# Install from Bioconductor (recommended)
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("spatialGE")

# Or install development version from GitHub
# remotes::install_github("ACSoupir-OC/spatialGE")
```

```{r load-library}
library(spatialGE)
```

## Loading Your Data

The core data structure in spatialGE is the `STlist` object, which stores:
- Count matrices for one or more samples
- Spot/cell coordinates
- Sample metadata
- Analysis results

### Example: Loading Visium Data

```{r load-data}
# For this vignette, we'll use the built-in test dataset
# In practice, replace with your data path
data_dir <- system.file("tests/testthat/data/tnbc_bassiouni", 
                        package = "spatialGE")

# Create STlist from Visium count matrices
st_obj <- STlist(
  data_dir = data_dir,
  data_type = "visium",
  min_spots = 100,
  min_counts = 500
)

# Inspect the object
st_obj
```

The `STlist` constructor automatically:
1. Reads count matrices from each sample directory
2. Loads spot coordinates from `spatial/tissue_positions_list.csv`
3. Filters spots and genes based on quality thresholds
4. Stores metadata for downstream analysis

### Supported Data Types

spatialGE supports multiple spatial transcriptomics platforms:

```{r data-types, eval=FALSE}
# Visium (10x Genomics)
STlist(data_dir = "path/to/visium", data_type = "visium")

# Xenium (10x Genomics)
STlist(data_dir = "path/to/xenium", data_type = "xenium")

# CosMx (Nanostring)
STlist(data_dir = "path/to/cosmx", data_type = "cosmx")

# Generic matrix format
STlist(data_dir = "path/to/matrices", 
       data_type = "generic",
       counts_file = "counts.mtx",
       coords_file = "coordinates.csv")
```

## Visualizing Gene Expression

Once your data is loaded, visualization is straightforward with `STplot`.

### Plot a Single Gene

```{r plot-gene, eval=FALSE}
# Visualize CD8A expression (T cell marker)
STplot(st_obj, 
       gene = "CD8A", 
       plot_type = "gene")
```

### Plot Multiple Genes

```{r plot-multi-gene, eval=FALSE}
# Visualize multiple genes side-by-side
genes_of_interest <- c("CD8A", "CD4", "MS4A1", "EPCAM")
STplot(st_obj, 
       gene = genes_of_interest, 
       plot_type = "gene",
       ncol = 2)
```

### Plot Clusters

After clustering, visualize the spatial patterns:

```{r plot-clusters, eval=FALSE}
# First, cluster the data (see next section)
st_obj <- STclust(st_obj, k = 8)

# Then visualize clusters
STplot(st_obj, plot_type = "clusters")
```

## Identifying Spatially Variable Genes

Spatial heterogeneity analysis (`SThet`) identifies genes with non-random spatial patterns using Moran's I spatial autocorrelation.

```{r sthet}
# Calculate spatial autocorrelation for all genes
# This may take a few minutes for large datasets
st_obj <- SThet(st_obj, 
                method = "moran",
                min_expr = 0.1)  # Filter lowly expressed genes

# Extract results
het_results <- st_obj@spatial_het

# View top spatially variable genes
head(het_results[order(-het_results$moranI), ])
```

The output includes:
- `moranI`: Moran's I statistic (higher = more spatially structured)
- `pval`: Statistical significance
- `fdr`: FDR-adjusted p-value

### Visualize Top Spatial Genes

```{r plot-spatial-genes, eval=FALSE}
# Get top 4 spatially variable genes
top_genes <- head(het_results$gene, 4)

# Visualize their spatial patterns
STplot(st_obj, gene = top_genes, plot_type = "gene", ncol = 2)
```

## Clustering and Spatial Domains

Identify spatially coherent tissue domains with `STclust`:

```{r stclust}
# Cluster spots into spatial domains
st_obj <- STclust(st_obj, 
                  k = 8,  # Number of clusters
                  method = "dynamicTreeCut")

# Access cluster assignments
clusters <- st_obj@clusters
head(clusters)
```

## What's Next?

You've completed the basic spatialGE workflow! For more advanced analyses, see:

- **Spatial Differential Expression**: `vignette("spatial_differential_expression")`
  - Compare gene expression between tissue domains
  - Account for spatial autocorrelation in statistical tests

- **Spatial Enrichment & Gradients**: `vignette("spatial_enrichment_gradients_smi")`
  - Gene set enrichment analysis with spatial context
  - Identify genes with directional expression patterns

- **Seurat Integration**: `vignette("seurat_integration")`
  - Convert between spatialGE and Seurat objects
  - Combine spatial and single-cell data

- **Troubleshooting**: `vignette("troubleshooting")`
  - Common issues and solutions
  - Performance optimization tips

## Need Help?

- **Documentation**: `browseVignettes("spatialGE")`
- **Function reference**: `?STlist`, `?STplot`, `?SThet`, `?STclust`
- **Issues**: Report bugs on [GitHub](https://github.com/ACSoupir-OC/spatialGE)
- **Citation**: `citation("spatialGE")`

## References

See `vignette("vignette_references")` for full bibliography.
