---
title: "Getting Started with spatialGE"
date: '`r format(Sys.Date())`'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{getting_started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r vig_settings, include=F}
knitr::opts_chunk$set(
  collapse=TRUE,
  comment="#>",
  fig.width=8, fig.height=5,
  eval=TRUE
)
options(rmarkdown.html_vignette.check_title=F)
```

This vignette provides a quick 5-minute tutorial to get started with the **`spatialGE`** package for spatial genomics analysis.

## Installation and Loading

Install the package from CRAN:

```{r install}
# install.packages('spatialGE')
```

Load the package:

```{r load}
library(spatialGE)
```

## Loading Example Data

The package includes TNBC (triple-negative breast cancer) example data from Visium spatial transcriptomics. Here we load one sample and create an STlist object:

```{r load_data}
# Get the path to the example data
data_path <- system.file("tests/testthat/data/tnbc_bassiouni", package="spatialGE")

# List available samples
samples <- list.dirs(data_path, full.names=TRUE, recursive=FALSE)
print(samples)

# Load clinical metadata
clin_file <- file.path(data_path, "bassiouni_clinical.csv")

# Create STlist with first sample
tnbc <- STlist(rnacounts=samples[1], samples=clin_file, cores=2)
print(tnbc)
```

## Step 1: Visualize Gene Expression with STplot

Plot expression of a single gene (e.g., *NDRG1*, a hypoxia marker):

```{r stplot_gene}
# Visualize gene expression
quilts <- STplot(tnbc, genes="NDRG1", samples="sample_094c", color_pal="BuRd")
quilts[[1]]
```

## Step 2: Calculate Spatial Heterogeneity with SThet

Compute Moran's I statistic to measure spatial clustering:

```{r sthet}
# Calculate spatial heterogeneity
tnbc <- SThet(tnbc, genes="NDRG1", method="moran", cores=2)

# View results
get_gene_meta(tnbc, sthet_only=TRUE)
```

## Step 3: Spatial Clustering with STclust

Perform unsupervised spatially-informed clustering:

```{r stclust}
# Perform clustering with weight w=0.05
tnbc <- STclust(tnbc, ks=5, ws=0.05, cores=2)

# Visualize clusters
cluster_plot <- STplot(tnbc, samples="sample_094c", ws=0.05, color_pal="highcontrast")
cluster_plot[[1]]
```

## Summary

That's it! You've now:

1. ✅ Loaded the `spatialGE` library
2. ✅ Loaded TNBC example data
3. ✅ Run `STplot` to visualize gene expression
4. ✅ Run `SThet` to calculate spatial heterogeneity (Moran's I)
5. ✅ Run `STclust` to perform spatial clustering
6. ✅ Created 3 plots total

For more advanced analyses, see the [basic_functions_vignette](basic_functions_vignette.html) or visit the [spatialGE documentation](https://fridleylab.github.io/spatialGE/).

## References

- Ospina J et al. (2022). "spatialGE: an R package for spatial genomics analysis."
