---
title: "Seurat Integration: Bidirectional Workflows"
date: '`r format(Sys.Date())`'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{seurat_integration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  eval = TRUE
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

## Overview

This vignette demonstrates how to integrate **spatialGE** with **Seurat** for seamless spatial transcriptomics workflows. The `spatialGE` package provides bidirectional conversion utilities that allow you to:

1. **Convert Seurat objects to STlist** - Run spatialGE analyses on Seurat-processed data
2. **Convert STlist to Seurat** - Leverage Seurat's visualization and downstream analysis tools
3. **Run spatialGE directly from Seurat** - One-step analysis workflows
4. **Transfer results back to Seurat** - Add cluster assignments, spatial statistics to Seurat metadata

## Installation

Make sure both packages are installed:

```{r install_packages, eval=FALSE}
# Install spatialGE
devtools::install_github("ACSoupir-OC/spatialGE")

# Install Seurat (if not already installed)
install.packages("Seurat")
```

Load the required libraries:

```{r load_libraries, message=FALSE, warning=FALSE}
library(spatialGE)
library(Seurat)
library(Matrix)
library(tibble)
```

## Creating Example Data

For this vignette, we'll create a synthetic STlist object to demonstrate the workflow:

```{r create_data, message=TRUE}
set.seed(42)

# Create synthetic count matrix (20 genes x 100 cells)
n_genes <- 20
n_cells <- 100

counts_mat <- Matrix::Matrix(
  matrix(rnbinom(n_genes * n_cells, mu = 5, size = 2), 
         nrow = n_genes, ncol = n_cells), 
  sparse = TRUE
)
rownames(counts_mat) <- paste0("Gene", 1:n_genes)
colnames(counts_mat) <- paste0("Cell", 1:n_cells)

# Add some marker gene names
rownames(counts_mat)[1:5] <- c("MLANA", "CD37", "TP53", "GAPDH", "ACTB")

# Create spatial coordinates
coords_df <- data.frame(
  barcode = paste0("Cell", 1:n_cells),
  xpos = runif(n_cells, 0, 1000),
  ypos = runif(n_cells, 0, 1000)
)

# Create STlist object
st_obj <- methods::new("STlist",
  counts = list(test_sample = counts_mat),
  spatial_meta = list(test_sample = coords_df),
  sample_meta = tibble::tibble(sample_name = "test_sample"),
  gene_meta = list()
)

cat(sprintf("Created STlist: %d genes x %d cells\n", n_genes, n_cells))
```

## Workflow 1: STlist to Seurat Conversion

Convert your STlist object to Seurat for visualization:

```{r stlist_to_seurat, message=TRUE}
# Convert STlist to Seurat object
seurat_obj <- as.Seurat.STlist(st_obj, verbose = FALSE)

# Inspect the Seurat object
cat(sprintf("Seurat object:\n"))
cat(sprintf("  Cells: %d\n", ncol(seurat_obj)))
cat(sprintf("  Features: %d\n", nrow(seurat_obj)))

# View metadata
head(seurat_obj@meta.data)
```

## Workflow 2: Seurat to STlist Conversion

Convert back from Seurat to STlist (round-trip validation):

```{r seurat_to_stlist, message=TRUE}
# Convert Seurat back to STlist
st_obj2 <- as.STlist.Seurat(seurat_obj, verbose = FALSE)

cat(sprintf("Round-trip validation:\n"))
cat(sprintf("  Original: %d genes x %d cells\n", 
            nrow(st_obj@counts[[1]]), ncol(st_obj@counts[[1]])))
cat(sprintf("  After round-trip: %d genes x %d cells\n", 
            nrow(st_obj2@counts[[1]]), ncol(st_obj2@counts[[1]])))
cat(sprintf("  Match: %s\n", 
            all(dim(st_obj@counts[[1]]) == dim(st_obj2@counts[[1]]))))
```

## Workflow 3: Add Results to Seurat Metadata

You can add custom analysis results to Seurat metadata:

```{r add_metadata, message=TRUE}
# Add cluster assignments
set.seed(42)
seurat_obj$spatialGE_cluster <- sample(c("A", "B", "C"), ncol(seurat_obj), replace = TRUE)

# Add spatial statistics
seurat_obj$moran_I <- runif(ncol(seurat_obj), 0, 0.5)

cat("Metadata columns added:\n")
head(seurat_obj@meta.data[, c("orig.ident", "spatialGE_cluster", "moran_I")])
```

## Workflow 4: Gene Set Integration

Convert spatialGE enrichment results for Seurat GSEA or AddModuleScore:

```{r geneset_integration, message=TRUE}
# Define pathway gene sets
pathway_list <- list(
  "Pathway1" = c("MLANA", "CD37", "TP53"),
  "Pathway2" = c("GAPDH", "ACTB")
)

# Convert to Seurat format
gene_sets <- spatialGE_to_seurat_genesets(
  list(test_sample = data.frame(
    geneset = names(pathway_list),
    genes = I(pathway_list),
    pval = c(0.01, 0.05),
    stringsAsFactors = FALSE
  )),
  pval.thr = 0.05,
  return.type = "list"
)

cat(sprintf("Generated %d gene sets\n", length(gene_sets)))
for (i in seq_along(gene_sets)) {
  cat(sprintf("  - %s: %d genes\n", names(gene_sets)[i], length(gene_sets[[i]])))
}
```

## Complete Example: End-to-End Workflow

Here's a complete workflow from start to finish:

```{r complete_workflow, eval=FALSE}
# ============================================================
# Complete workflow from STlist to Seurat and back
# ============================================================

# Step 1: Create or load STlist object
# st_obj <- STlist(rnacounts = count_files, spotcoords = coord_files)

# Step 2: Run spatialGE analysis
st_obj <- SThet(st_obj, genes = c("MLANA", "CD37", "TP53"))
st_obj <- STclust(st_obj, k = 3)

# Step 3: Convert to Seurat for visualization
seurat_obj <- as.Seurat.STlist(st_obj, verbose = TRUE)

# Step 4: Add results to Seurat metadata
seurat_obj <- add_spatialGE_to_seurat(seurat_obj, st_obj, result_type = "clusters")

# Step 5: Visualize in Seurat
# SpatialDimPlot(seurat_obj, group.by = "spatialGE_cluster")
# FeaturePlot(seurat_obj, features = c("MLANA", "CD37"))

# Step 6: Save results
# saveRDS(seurat_obj, "seurat_with_spatialge_results.rds")
```

## Troubleshooting

### Issue: No spatial coordinates found

Check if your object has spatial information:

```{r troubleshooting_coords, eval=FALSE}
# For Seurat objects, check images slot
names(seurat_obj@images)

# For STlist objects, check spatial_meta
names(st_obj@spatial_meta)
```

### Issue: Memory errors with large objects

Process samples individually:

```{r troubleshooting_memory, eval=FALSE}
# Process one sample at a time
for (sample in names(st_obj@counts)) {
  st_single <- st_obj[, sample]
  st_single <- SThet(st_single, genes = genes_of_interest)
}
```

## Best Practices

1. **Always validate after conversion**
   ```{r best_practice_validate, eval=FALSE}
   st_obj <- as.STlist.Seurat(seurat_obj)
   print(st_obj)
   summary(st_obj)
   ```

2. **Preserve original objects**
   ```{r best_practice_backup, eval=FALSE}
   seurat_original <- seurat_obj  # Keep backup
   seurat_obj <- as.Seurat.STlist(st_obj)  # Overwrite carefully
   ```

3. **Use raw counts for spatialGE**
   ```{r best_practice_assay, eval=FALSE}
   # Check available assays
   Seurat::Assays(seurat_obj)
   
   # Use raw counts
   st_obj <- as.STlist.Seurat(seurat_obj, assay = "RNA", slot = "counts")
   ```

## See Also

- `vignette("basic_functions_vignette", package = "spatialGE")` - Core spatialGE functions
- Seurat website: https://satijalab.org/seurat/

## References

- Hao Y, et al. (2024). "Seurat v5." _Nature Biotechnology_.
- Stuart T, et al. (2019). "Single-Cell Data Integration." _Cell_.
