---
title: "Seurat Integration: Bidirectional Workflows"
date: '`r format(Sys.Date())`'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{seurat_integration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  eval = TRUE
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

## Overview

This vignette demonstrates how to integrate **spatialGE** with **Seurat** for seamless spatial transcriptomics workflows. The `spatialGE` package provides bidirectional conversion utilities that allow you to:

1. **Convert Seurat objects to STlist** - Run spatialGE analyses on Seurat-processed data
2. **Convert STlist to Seurat** - Leverage Seurat's visualization and downstream analysis tools  
3. **Run spatialGE directly from Seurat** - One-step analysis workflows
4. **Transfer results back to Seurat** - Add cluster assignments, spatial statistics to Seurat metadata

## Installation

Make sure both packages are installed:

```{r install_packages, eval=FALSE}
# Install spatialGE
devtools::install_github("ACSoupir-OC/spatialGE")

# Install Seurat (if not already installed)
install.packages("Seurat")
```

Load the required libraries:

```{r load_libraries, message=FALSE, warning=FALSE}
library(spatialGE)
library(Seurat)
library(Matrix)
library(tibble)
```

## Workflow 1: Creating Test Data

For this vignette, we'll create a synthetic STlist object to demonstrate the workflow:

```{r create_test_data, message=TRUE}
set.seed(42)

# Create synthetic count matrix (20 genes x 100 cells for vignette demo)
# Using smaller size to avoid C stack issues during vignette build
n_genes <- 20
n_cells <- 100

counts_mat <- Matrix(matrix(rnbinom(n_genes * n_cells, mu = 5, size = 2), 
                            nrow = n_genes, ncol = n_cells), sparse = TRUE)
rownames(counts_mat) <- paste0("Gene", 1:n_genes)
colnames(counts_mat) <- paste0("Cell", 1:n_cells)

# Add some known marker genes
rownames(counts_mat)[1:10] <- c("MLANA", "CD37", "TP53", "COL1A1", "ACTB", 
                                 "GAPDH", "VIM", "CDH1", "FN1", "SNAI1")

# Update genes_of_interest for smaller dataset
genes_of_interest <- c("MLANA", "CD37", "TP53", "GAPDH", "ACTB")

# Create spatial coordinates
coords_df <- data.frame(
  barcode = paste0("Cell", 1:n_cells),
  xpos = runif(n_cells, 0, 1000),
  ypos = runif(n_cells, 0, 1000)
)

# Create STlist object
st_obj <- methods::new("STlist",
  counts = list(test_sample = counts_mat),
  spatial_meta = list(test_sample = coords_df),
  sample_meta = tibble(sample_name = "test_sample"),
  gene_meta = list()
)

# Transform data (required before analysis)
st_obj <- transform_data(st_obj)

cat(sprintf("Created STlist:\n"))
cat(sprintf("  Genes: %d\n", nrow(st_obj@counts[[1]])))
cat(sprintf("  Cells: %d\n", ncol(st_obj@counts[[1]])))
cat(sprintf("  Samples: %d\n", length(st_obj@counts)))
```

## Workflow 2: STlist → Seurat Conversion

Convert your STlist object to Seurat for visualization:

```{r stlist_to_seurat, message=TRUE}
# Convert STlist to Seurat object
seurat_obj <- as.Seurat.STlist(st_obj, verbose = FALSE)

# Inspect the Seurat object
cat(sprintf("Seurat object created:\n"))
cat(sprintf("  Class: %s\n", class(seurat_obj)[1]))
cat(sprintf("  Cells: %d\n", ncol(seurat_obj)))
cat(sprintf("  Features: %d\n", nrow(seurat_obj)))
cat(sprintf("  Assays: %s\n", paste(names(seurat_obj@assays), collapse = ", ")))

# Check metadata structure
cat("\nMetadata columns:\n")
cat(paste(colnames(seurat_obj@meta.data), collapse = ", "), "\n")

# View first few cells
cat("\nFirst 6 cells:\n")
head(seurat_obj@meta.data)
```

## Workflow 3: Seurat → STlist Conversion

Convert back from Seurat to STlist (round-trip validation):

```{r seurat_to_stlist, message=TRUE}
# Convert Seurat back to STlist
st_obj2 <- as.STlist.Seurat(seurat_obj, verbose = FALSE)

cat(sprintf("Round-trip validation:\n"))
cat(sprintf("  Original: %d genes × %d cells\n", 
            nrow(st_obj@counts[[1]]), ncol(st_obj@counts[[1]])))
cat(sprintf("  After round-trip: %d genes × %d cells\n", 
            nrow(st_obj2@counts[[1]]), ncol(st_obj2@counts[[1]])))
cat(sprintf("  Match: %s\n", 
            all(dim(st_obj@counts[[1]]) == dim(st_obj2@counts[[1]]))))
```

## Workflow 4: Run spatialGE Analysis

Run spatial heterogeneity analysis on the STlist object:

```{r run_sthet, message=TRUE}
# Run spatial heterogeneity analysis
genes_of_interest <- c("MLANA", "CD37", "TP53", "GAPDH", "ACTB")
st_obj <- SThet(st_obj, genes = genes_of_interest, verbose = FALSE)

# View results
cat("SThet results:\n")
print(st_obj@gene_meta[[1]][, c("gene", "moranI", "gearyC", "pval_moran")])
```

## Workflow 5: Add Results to Seurat Metadata

Transfer spatialGE analysis results back to Seurat:

```{r add_to_seurat, message=TRUE}
# Add Moran's I values to Seurat feature metadata
seurat_obj <- add_spatialGE_to_seurat(
  seurat_obj, st_obj,
  result_type = "moran",
  verbose = FALSE
)

cat("Moran's I values added to Seurat:\n")
head(seurat_obj@assays$RNA@meta.features)

# Run clustering
st_obj <- STclust(st_obj, k = 3, verbose = FALSE)

# Add cluster assignments
seurat_obj <- add_spatialGE_to_seurat(
  seurat_obj, st_obj,
  result_type = "clusters",
  verbose = FALSE
)

cat("\nCluster assignments:\n")
table(seurat_obj$spatialGE_cluster)
```

## Workflow 6: Direct Analysis with spatialGE_from_seurat()

For quick analyses, use the wrapper function:

```{r direct_analysis, message=TRUE}
# One-step heterogeneity analysis directly from Seurat
result <- spatialGE_from_seurat(
  seurat_obj,
  analysis = "SThet",
  genes = genes_of_interest,
  assay = "RNA",
  verbose = FALSE
)

cat("Direct analysis results:\n")
print(result@gene_meta[[1]][1:3, c("gene", "moranI", "pval_moran")])
```

## Workflow 7: Gene Set Integration

Convert spatialGE enrichment results for Seurat GSEA:

```{r geneset_integration, message=TRUE}
# Define pathway gene sets
pathway_list <- list(
  "Melanocyte" = c("MLANA", "PMEL", "TYR", "DCT"),
  "Immune" = c("CD37", "PTPRC", "CD3D", "CD8A"),
  "EMT" = c("VIM", "CDH2", "FN1", "SNAI1")
)

# Run enrichment analysis
enrich_result <- STenrich(st_obj, gene_sets = pathway_list, verbose = FALSE)

cat("Enrichment results:\n")
print(enrich_result[[1]][, c("geneset", "pval", "genes")])

# Convert to Seurat gene sets (p < 0.05)
gene_sets <- spatialGE_to_seurat_genesets(
  enrich_result,
  pval.thr = 0.05,
  return.type = "list"
)

cat(sprintf("\nGenerated %d gene sets (p < 0.05)\n", length(gene_sets)))

# Add module scores to Seurat
if (length(gene_sets) > 0) {
  seurat_obj <- AddModuleScore(
    seurat_obj,
    features = gene_sets,
    name = "spatialGE_"
  )
  
  cat("\nModule scores added:\n")
  score_cols <- grep("spatialGE_", colnames(seurat_obj@meta.data), value = TRUE)
  head(seurat_obj@meta.data[, score_cols])
}
```

## Complete Example: End-to-End Workflow

```{r complete_workflow, eval=FALSE}
# ============================================================
# Complete workflow from Seurat to spatialGE and back
# ============================================================

# Step 1: Start with Seurat object
# seurat_obj <- Load10X_Spatial(data.dir = "visium_data/")

# Step 2: Convert to STlist
st_obj <- as.STlist.Seurat(seurat_obj, verbose = TRUE)

# Step 3: Run spatial heterogeneity analysis
genes <- c("MLANA", "CD37", "TP53", "GAPDH", "ACTB")
st_obj <- SThet(st_obj, genes = genes)

# Step 4: Run spatial clustering
st_obj <- STclust(st_obj, k = 5)

# Step 5: Convert back to Seurat
seurat_obj <- as.Seurat.STlist(st_obj, verbose = TRUE)

# Step 6: Add results to Seurat metadata
seurat_obj <- add_spatialGE_to_seurat(seurat_obj, st_obj, result_type = "clusters")
seurat_obj <- add_spatialGE_to_seurat(seurat_obj, st_obj, result_type = "moran")

# Step 7: Visualize in Seurat
# SpatialDimPlot(seurat_obj, group.by = "spatialGE_cluster")
# FeaturePlot(seurat_obj, features = c("MLANA", "CD37"))

# Step 8: Save results
# saveRDS(seurat_obj, "seurat_with_spatialge_results.rds")
```

## Troubleshooting

### Issue: No spatial coordinates found

```{r troubleshooting_coords, eval=FALSE}
# Check for spatial data
names(seurat_obj@images)

# If empty, coordinates won't be extracted
```

### Issue: Memory errors with large objects

```{r troubleshooting_memory, eval=FALSE}
# Process samples one at a time
for (sample in names(st_obj@counts)) {
  st_single <- st_obj[, sample]
  st_single <- SThet(st_single, genes = genes_of_interest)
}
```

## Best Practices

1. **Validate after conversion**
   ```{r best_practice_validate, eval=FALSE}
   st_obj <- as.STlist.Seurat(seurat_obj)
   print(st_obj)
   ```

2. **Preserve original objects**
   ```{r best_practice_backup, eval=FALSE}
   seurat_original <- seurat_obj  # Keep backup
   ```

3. **Use raw counts for spatialGE**
   ```{r best_practice_assay, eval=FALSE}
   st_obj <- as.STlist.Seurat(seurat_obj, assay = "RNA", slot = "counts")
   ```

## See Also

- `vignette("basic_functions_vignette", package = "spatialGE")` - Core spatialGE functions
- Seurat website: https://satijalab.org/seurat/

## References

- Hao Y, et al. (2024). "Seurat v5." _Nature Biotechnology_.
- Stuart T, et al. (2019). "Single-Cell Data Integration." _Cell_.
