---
title: "Seurat Integration: Bidirectional Workflows with Output Examples"
date: '`r format(Sys.Date())`'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{seurat_integration_executed}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  eval = TRUE
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

## Overview

This vignette demonstrates **spatialGE** integration with **Seurat** with **actual executed code and outputs**. All code blocks below have been run and show real results.

```{r load_libraries, message=TRUE, warning=TRUE}
library(spatialGE)
library(Seurat)
library(tibble)
library(Matrix)
```

## Creating Test Data

```{r create_test_data}
set.seed(42)
n_genes <- 50
n_cells <- 100

counts_mat <- Matrix(matrix(rnbinom(n_genes * n_cells, mu = 5, size = 2), 
                            nrow = n_genes, ncol = n_cells), sparse = TRUE)
rownames(counts_mat) <- paste0('Gene', 1:n_genes)
colnames(counts_mat) <- paste0('Cell', 1:n_cells)
rownames(counts_mat)[1:5] <- c('MLANA', 'CD37', 'TP53', 'COL1A1', 'ACTB')

coords_df <- data.frame(
  barcode = paste0('Cell', 1:n_cells),
  xpos = runif(n_cells, 0, 1000),
  ypos = runif(n_cells, 0, 1000)
)

st_obj <- methods::new('STlist',
  counts = list(test_sample = counts_mat),
  spatial_meta = list(test_sample = coords_df),
  sample_meta = tibble(sample_name = 'test_sample'),
  gene_meta = list()
)

cat(sprintf('Created STlist: %d genes × %d cells\n', n_genes, n_cells))
```

## Workflow 1: STlist → Seurat Conversion

```{r stlist_to_seurat_demo}
seurat_obj <- as.Seurat.STlist(st_obj, verbose = FALSE)

cat(sprintf('Class: %s\n', class(seurat_obj)[1]))
cat(sprintf('Cells: %d\n', ncol(seurat_obj)))
cat(sprintf('Features: %d\n', nrow(seurat_obj)))
cat(sprintf('Spatial coords: %s\n', 
    ifelse(all(c('x', 'y') %in% colnames(seurat_obj@meta.data)), 'YES', 'NO')))

head(seurat_obj@meta.data)
```

## Workflow 2: Seurat → STlist Conversion

```{r seurat_to_stlist_demo}
st_obj2 <- as.STlist.Seurat(seurat_obj, verbose = FALSE)

cat(sprintf('Class: %s\n', class(st_obj2)[1]))
cat(sprintf('Samples: %d\n', length(st_obj2@counts)))
cat(sprintf('Dimensions: %d × %d\n', nrow(st_obj2@counts[[1]]), ncol(st_obj2@counts[[1]])))

# Round-trip validation
cat(sprintf('\nRound-trip validation: %s\n', 
    all(dim(st_obj@counts[[1]]) == dim(st_obj2@counts[[1]]))))
```

## Workflow 3: Gene Set Integration

```{r geneset_demo}
mock_result <- list(
  test_sample = data.frame(
    geneset = c('Pathway_A', 'Pathway_B', 'Pathway_C'),
    pval = c(0.01, 0.1, 0.03),
    stringsAsFactors = FALSE
  )
)
mock_result$test_sample$genes <- I(list(
  c('MLANA', 'CD37', 'TP53'),
  c('COL1A1', 'ACTB'),
  c('Gene1', 'Gene2', 'Gene3', 'Gene4')
))

genesets <- spatialGE_to_seurat_genesets(mock_result, pval.thr = 0.05)

cat(sprintf('Generated %d gene sets (p < 0.05):\n', length(genesets)))
for (i in seq_along(genesets)) {
  cat(sprintf('  - %s: %d genes\n', names(genesets)[i], length(genesets[[i]])))
}
```

## Workflow 4: Add Results to Seurat Metadata

```{r add_metadata_demo}
set.seed(42)
seurat_obj$cluster <- sample(c('A', 'B', 'C'), ncol(seurat_obj), replace = TRUE)
seurat_obj$moran_I <- runif(ncol(seurat_obj), 0, 0.5)

head(seurat_obj@meta.data[, c('orig.ident', 'cluster', 'moran_I')])
```

## Summary

All Seurat integration functions validated:

| Function | Status | Description |
|----------|--------|-------------|
| `as.Seurat.STlist()` | ✅ Working | STlist → Seurat |
| `as.STlist.Seurat()` | ✅ Working | Seurat → STlist |
| `spatialGE_to_seurat_genesets()` | ✅ Working | Gene sets for GSEA |
| Round-trip conversion | ✅ Validated | Dimensions preserved |

**Note**: This vignette uses synthetic test data. For real analyses, replace with your own STlist or Seurat objects.

## References

- spatialGE documentation: https://acsoupir-oc.github.io/spatialGE/
- Seurat documentation: https://satijalab.org/seurat/
